---
title: "Transmission Type Fuel Effieciency Analysis"
output: pdf_document
date: '2014-10-26'
---

### Executive Summary
Using the Motor Trends data this report endeavours to find out whether automatic or manual transmission is better for MPG. This report will further more attempt to quantify the MPG difference between automatic and manual transmissions. In an effort to keep the report concise most of the R code is not shown, please feel free to visit this [GitHub Repository](https://github.com/schonken/Regression_Models_Project) for the complete R Markdown code.
 
### Data Dictionary
The Motor Trends data consists of 32 observations spanning 11 variables. For a full description of the data definition please refer to the [Motor Trends Data Documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).


Field | Definition | Field | Definition
----- | ----- | ----- | ----- 
mpg | Miles/(US) gallon | cyl | Number of cylinders
disp | Displacement (cu.in.) | hp | Gross horsepower
drat | Rear axle ratio | wt | Weight (lb/1000)
qsec | 1/4 mile time in seconds | vs | V/S (0 = V-engine, 1 = straight engine)
am | Transmission (0 = automatic, 1 = manual) | gear | Number of forward gears
carb | Number of carburetors

### Exploring the Data
We start with a violin plot of ```MPG vs Transmission``` to explore the relatioship between transmission type and MPG. Next we run ```summary()``` on a few of the more interesting fields  in the data to get a feel of what the data is telling us. 

```{r Init, echo=FALSE, message=FALSE, fig.height=3.25}
library(car)


cars <- mtcars

require(ggplot2)
ggplot(data=cars, aes(x=factor(am), y=mpg, fill=factor(am))) +
  geom_violin() +
  ylab('Miles/(US) gallon') + 
  xlab('Transmission (0 = automatic, 1 = manual)') + 
  ggtitle('MPG vs Transmission')

summary(data.frame(cars$mpg, cars$cyl, cars$disp, cars$hp, cars$drat, cars$wt, cars$qsec, cars$carb))

mpg.mean <- c(mean(cars[cars$am==0, 'mpg']), mean(cars[cars$am==1, 'mpg']))
```

We find that the mean MPG for our Automatic transmissions is `r mpg.mean[1]` and for Manual transmissions is `r mpg.mean[2]`. This is a mean MPG difference of `r mpg.mean[2] - mpg.mean[1]`  with the Automatic transmission being less efficient. This extremely simple comparison is just the beginning of our journey.

### Regression Modelling
We iteratively run linear regressions starting with ```summary(lm(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef```, in each subsequent run we remove the entry (field) with the largest ```Pr(>|t|)``` score until we have only one left, namely ```wt```. Next we use the ```anova()``` function to find the cut-off point in terms regression variables. We find that including the ```gear``` is not beneficial to our regression model and when it is not specified the Transmission type ```am``` scrapes in just under 5%.  

```{r Anova, echo=FALSE}
# The following block of commented code was used to find the most important fields for our regression
# summary(lm(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ disp + hp + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ hp + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ wt + qsec + am, data = cars))$coef
# summary(lm(mpg ~ wt + qsec, data = cars))$coef
# summary(lm(mpg ~ wt, data = cars))$coef

fitA <- lm(mpg ~ wt, data = cars)
fitB <- lm(mpg ~ wt + qsec, data = cars)
fitC <- lm(mpg ~ wt + qsec + am, data = cars)
fitD <- lm(mpg ~ wt + qsec + am + gear, data = cars)

fitAtoC <- anova(fitA, fitB, fitC)[6]
fitAtoD <- anova(fitA, fitB, fitC, fitD)[6]
```
Field | Pr(>F) for (mpg ~ wt, qsec, am) | Pr(>F) for (mpg ~ wt, qsec, am, gear)
----- | ----- | -----
wt | `r fitAtoC[1,1]` | `r fitAtoD[1,1]`
qsec | `r fitAtoC[2,1]` | `r fitAtoD[2,1]`
am | `r fitAtoC[3,1]` | `r fitAtoD[3,1]`
gear | | `r fitAtoD[4,1]`

The fields ```wt```, ```qsec``` and ```am``` feel like the right set of variables to be looking at but a linear coefficient against ```qsec``` seems odd. One would expect diminishing returns for the fuel consumed as the ```qsec``` gets smaller (i.e. increased acceleration). In an effort to model these diminishing returns I introduce a new variable named ```qsec.2.I``` which represents the inverse of the square of ```qsec``` ($1/qsec^2$). The following two tables detail the coefficients of the respective linear regression models. We see that using the variable ```qsec.2.I``` as opposed to ```qsec``` yields a tighter fit to our data as evidenced by every single ```Pr(>|t|)``` being almost halved.

```{r lm, echo=FALSE}
cars <- cbind(cars, qsec.2.I = 1/(cars$qsec*cars$qsec))

lm.qsec <- summary(lm(mpg ~ wt + qsec + am, data = cars))$coef
lm.qsec.2.I <- summary(lm(mpg ~ wt + qsec.2.I + am, data = cars))$coef

lm.qsec
lm.qsec.2.I

lm.qsec <- summary(lm(mpg ~ wt + qsec + am, data = cars))
lm.qsec.2.I <- summary(lm(mpg ~ wt + qsec.2.I + am, data = cars))
```

### Interpreting the Coefficients
Using the ```qsec``` based regression we arrive at following model. For each additional ```wt``` (1000 lb) our MPG decreases by `r -lm.qsec$coef[2,1]`. For each additional second needed to complete ```qsec``` (1/4 mile time in seconds) our MPG increases by `r lm.qsec$coef[3,1]`. Lastly, Manual transmissions yield a `r lm.qsec$coef[4,1]` MPG benefit (increase) over Automatic transmissions. 

### Confidence
The $R^2$ for the ```qsec``` and ```qsec.2.I``` regression models are `r lm.qsec$r.sq` and `r lm.qsec.2.I$r.sq` respectively. The $R^2$ for the ```qsec.2.I``` based regression model is marginally stronger (by `r lm.qsec.2.I$r.sq - lm.qsec$r.sq`) and with this model explaining `r lm.qsec.2.I$r.sq` of the variability it is a highly predictive model. We opt to use the ```qsec.2.I``` based regression model going forward.

### Residuals
```{r residplot, fig.height=3.5, echo=FALSE}
fit <- lm(mpg ~ wt + qsec.2.I + am, data = cars)
ggplot(cars, aes(x=predict(fit), y=resid(fit))) +
  geom_point(shape=1) + geom_smooth(method=lm) +
  ylab('Residuals') + xlab('Predicted') + ggtitle('Residual vs Predicted')
```

The Residuals vs. Predicted look good for this model and I believe we can in good conscience use it.

### Conclusion
Using our ```qsec.2.I``` based regression model we find that Manual transmissions yield better MPG in the amount of `r lm.qsec.2.I$coef[4,1]` MPG.

