---
title: "Transmission Type fuel effieciency Analysis"
author: "Francois Schonken"
date: "26/10/2014"
output: pdf_document
---

### Executive Summary
Using the Motor Trends data this report endeavours to find out whether automatic or manual transmission is better for MPG? This report will further more attempt to quantify the MPG difference between automatic and manual transmissions. 

In an effort to keep the report consise most R code is not show, please feel free to visit this [GitHub Repo](https://github.com/schonken/Regression_Models_Project) for the complete R Markdown code.

### Data Dictionary
The Motor Trends data consists of 32 observations. For a full description of the data definition please refer to the [Motor Trends Data Documentation](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).


Field | Definition | Field | Definition
----- | ----- | ----- | ----- 
mpg | Miles/(US) gallon | cyl | Number of cylinders
disp | Displacement (cu.in.) | hp | Gross horsepower
drat | Rear axle ratio | wt | Weight (lb/1000)
qsec | 1/4 mile time in seconds | vs | V/S (0 = V-engine, 1 = straight engine)
am | Transmission (0 = automatic, 1 = manual) | gear | Number of forward gears
carb | Number of carburetors

### Exploring the Motor Trend data
Let's start by getting a feel for the dataset by running ```summuary()``` on a few of the more interesting fields followed by a violin plot to get an idea of the role transmission plays in MPG. 
```{r Init, echo=FALSE, message=FALSE, fig.height=3.5}
library(car)


cars <- mtcars
summary(data.frame(cars$mpg, cars$cyl, cars$disp, cars$hp, cars$drat, cars$wt, cars$qsec, cars$carb))

require(ggplot2)
ggplot(data=cars, aes(x=factor(am), y=mpg, fill=factor(am))) +
  geom_violin() +
  ylab('Miles/(US) gallon') + 
  xlab('Transmission (0 = automatic, 1 = manual)') + 
  ggtitle('MPG vs Transmission')

mpg.mean <- c(mean(cars[cars$am==0, 'mpg']), mean(cars[cars$am==1, 'mpg']))
```

We find that the mean MPG for our Automatic transmisions is `r mpg.mean[1]` and `r mpg.mean[2]` for our Manual transmissions giving us a `r mpg.mean[2] - mpg.mean[1]` mean MPG difference with the Automatic transmission being less effiecient. This extremely simple comparison is just the beginning of our journey.

### Regression Modeling
We iteratively run linear regressions starting with ```summary(lm(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef```, in each subsequent run we remove the entry (field) with the largest ```Pr(>|t|)``` score until we have only one left, namely ```wt```. Next we use the ```anova()``` function to find the cut-off point in terms regression variables. We find that including the ```gear``` is not beneficial to our regression model and when it is not specified the Transmission type ```am``` scrapes in just under 5%.  

```{r Anova, echo=FALSE}
# The following block of commented code was used to find the most important fields for our regression
# summary(lm(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + vs + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + am + gear + carb, data = cars))$coef
# summary(lm(mpg ~ disp + hp + drat + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ disp + hp + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ hp + wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ wt + qsec + am + gear, data = cars))$coef
# summary(lm(mpg ~ wt + qsec + am, data = cars))$coef
# summary(lm(mpg ~ wt + qsec, data = cars))$coef
# summary(lm(mpg ~ wt, data = cars))$coef

fitA <- lm(mpg ~ wt, data = cars)
fitB <- lm(mpg ~ wt + qsec, data = cars)
fitC <- lm(mpg ~ wt + qsec + am, data = cars)
fitD <- lm(mpg ~ wt + qsec + am + gear, data = cars)

fitAtoC <- anova(fitA, fitB, fitC)[6]
fitAtoD <- anova(fitA, fitB, fitC, fitD)[6]
```
Field | Pr(>F) for (mpg ~ wt, qsec, am) | Pr(>F) for (mpg ~ wt, qsec, am, gear)
----- | ----- | -----
wt | `r fitAtoC[1,1]` | `r fitAtoD[1,1]`
qsec | `r fitAtoC[2,1]` | `r fitAtoD[2,1]`
am | `r fitAtoC[3,1]` | `r fitAtoD[3,1]`
gear | | `r fitAtoD[4,1]`

The fields ```wt```, ```qsec``` and ```am``` feel like the right set of variables to be looking at but a linear coeficient against ```qsec``` seems odd. One would expect diminishing returns for the fuel consumed as the ```qsec``` gets smaller (i.e. increased accelaration). In an effort to model these diminishing returns I introduce a new variable named ```qsec.2.I``` which represents the inverse of the square of ```qsec``` ($1/qsec^2$). The following two tables detail the coeficients of the respective linear regression models. We see that using the variable ```qsec.2.I``` as opposed to ```qsec``` yields a tighter fit to our data as evidenced by every single ```Pr(>|t|)``` being almost halved.

```{r lm, echo=FALSE}
cars <- cbind(cars, qsec.2.I = 1/(cars$qsec*cars$qsec))

lm.qsec <- summary(lm(mpg ~ wt + qsec + am, data = cars))$coef
lm.qsec.2.I <- summary(lm(mpg ~ wt + qsec.2.I + am, data = cars))$coef

lm.qsec
lm.qsec.2.I

lm.qsec <- summary(lm(mpg ~ wt + qsec + am, data = cars))
lm.qsec.2.I <- summary(lm(mpg ~ wt + qsec.2.I + am, data = cars))
```

Using the ```qsec``` based regression we arive at a model that reads for each ```wt``` (1000 lb) increase MPG decreases by `r lm.qsec$coef[2,1]`. For each additional second in ```qsec``` (1/4 mile time in seconds) MPG increases by `r lm.qsec$coef[3,1]` and Manual transmissions yield a `r lm.qsec$coef[4,1]` MPG benefit over Automatic transmissions. 

### Confidence
The $R^2$ for the ```qsec``` and ```qsec.2.I``` regression models are `r lm.qsec$r.sq` and `r lm.qsec.2.I$r.sq` respectively. The $R^2$ for the ```qsec.2.I``` based regression model is marinally stronger (by `r lm.qsec.2.I$r.sq - lm.qsec$r.sq`) and with this model explaining `r lm.qsec.2.I$r.sq` of the variability this is a highly predictive model. We opt to use the ```qsec.2.I``` based regression model going forward.

### Residuals
```{r residplot, fig.height=3.5, echo=FALSE}
fit <- lm(mpg ~ wt + qsec.2.I + am, data = cars)
ggplot(cars, aes(x=predict(fit), y=resid(fit))) +
  geom_point(shape=1) + geom_smooth(method=lm) +
  ylab('Residuals') + xlab('Predicted') + ggtitle('Residual vs Predicted')
```

The Residuals vs. Predicted look good for this model and I believe we can in good concience use it.

### Conclusion
Using our ```qsec.2.I``` based regression model we find that Manual transmissions yeild better MPG in the amount of `r lm.qsec.2.I$coef[4,1]` MPG.

